The fundamental shape of dynamic experiences is preserved in episodic memory

### 2018-07-06 boneyard.tex
checking in intro changes

>>>>

Another major difference between traditional trial-based memory paradigms and real-world memory concerns the subjective experience of the participant.  For example, consider an emotionally salient event in your life (e.g.\ marriage, the birth of a child, death of a loved one, etc.), or even a movie that you were especially impacted by.  No list-learning paradigm (or similar) can hope to capture the nuance and depth of such experiences.  Our everyday experiences feel ``important'' to us, whereas the words we study on a random word list do not.

### 2018-07-06 main.tex
checking in intro changes

Another major difference between traditional trial-based memory paradigms and real-world memory concerns the subjective experience of the participant.  For example, consider an emotionally salient event in your life (e.g.\ marriage, the birth of a child, death of a loved one, etc.), or even a movie that you were especially impacted by.  No list-learning paradigm (or similar) can hope to capture the nuance and depth of such experiences.  Our everyday experiences feel ``important'' to us, whereas the words we study on a random word list do not.  One component of naturalistic experiences that enhances their impact concerns the temporal structure with which they unfold.  For example, our experiences in the real world are necessarily autocorrelated in space and time on short timescales.  Further, or experiences are often characterized by longer timescale correlations that reflect the impact of our past actions and observations.  These temporal dynamics are not typically present in traditional memory studies, but are important if we wish to understand how our memory systems remember our everyday experiences.

Despite the fact that information flows continuously through our sensory systems, prior research suggests that our memories are partitioned and centered around particular ``events'' \citep{Radv12, RadvZack11, BrunEtal18}. Short timescale autocorrelations in the contents of our experiences may cause information to become associatively linked in memory, while sharp contrasts (i.e. ``event boundaries'') may segment continuous experiences into events \citep{HeusEtal18, BrunEtal18, Radv12}. There is compelling evidence that event boundaries can influence memory organization in trial-based recognition \citep{HeusEtal18}, navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, the extent to which people naturally (verbally) recount their experiences in segments centered around particular events has not been explored.

To study the temporal and semantic aspects of memory for naturalistic experiences, we analyzed an open dataset in which participants view and verbally recall an episode of the BBC series \textit{Sherlock}  \citep{ChenEtal17}. We developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17} to capture the temporal evolution of rich semantic structure present in naturalistic experiences. We then use this model to characterize verbally recalled episodic memories for naturalistic experiences. We hypothesized that memories for naturalistic experiences are 1) segmented and centered around particular events and 2) less about recalling specific details and more about preserving the overall ``shape'' of an experience.

>>>>

One hallmark feature of naturalistic experiences is the temporal structure with which they unfold.  For example, our experiences in the real world are typically autocorrelated in space and time on short timescales.  In other words, the contents of our experiences from moment to moment are similar, predictable and (often) change slowly over time.  Further, or experiences are often characterized by longer timescale correlations that reflect the impact of our past actions and observations. For instance, distant memories can influence current perception and actions. These temporal dynamics are not typically present in traditional memory studies, but are likely important if we wish to understand how our memory systems remember our everyday experiences.

Despite the continuous nature of naturalistic experiences, prior research suggests that our memories are segmented and centered around particular ``events'' \citep{Radv12, RadvZack11, BrunEtal18}. Short timescale autocorrelations in the contents of our experiences may cause information to become associatively linked in memory, while sharp contrasts (i.e. ``event boundaries'') may segment continuous experiences into events \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization in trial-based recognition \citep{HeusEtal18}, navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, the extent to which people naturally recount their experiences as ``events'' has not been explored.

 Here, we propose that naturalistic experiences have a unique geometric `shape' formed by correlational structure in perceptual/conceptual aspects of the stimulus. For instance, if a particular location is revisited more than once during a naturalistic experience (e.g. watching a movie), the perceptual features will be more similar during those timepoints as compared to other timepoints in distinct locations.  Similarly, if two actors hold a conversation spanning multiple locations, conceptual espects of the stimulus will be more similar during those timepoints as compared to other timepoints.  Together, both perceptual and conceptual features contribute to the overall shape of a naturalistic stimulus.  As a consequence of this structure, we hypothesize that episodic memories for those same experiences should mirror this shape. While the specific details and words recalled may be different from person to person, we expect that the overall shape of the experience will be preserved.

To study the temporal and semantic aspects of memory for naturalistic experiences, we analyzed an open dataset in which participants view and verbally recall an episode of the BBC series \textit{Sherlock}  \citep{ChenEtal17}. We developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17} to capture the temporal evolution of rich semantic structure present in naturalistic experiences. We then use this model to characterize verbally recalled episodic memories for naturalistic experiences.
